---
title: "Parental leave across OECD countries"
author: "Olga Mierzwa-Sulima"
date: 2019-05-03
categories: ["R"]
tags: ["R", "plotly", "parental-leave"]
---

## A quick look into paid leave available to moms

I was recently researching maternity leave options in Poland. In simplification as a mother in Poland you can get:

* a year of a paid leave at a rate 80% (20 weeks of maternity leave and 32 of parental leave)
* half a year (20 weeks maternity leave + 6 weeks parental leave) paid 100% and another half 60% (26 weeks of parental leave). 

You can use your parental leave straight after maternity leave or in chunks, split with the baby father or combine it with work. I won't go into too many details of the Polish social security system.
I was curious about how this compares to what other countries have to offer. I found out that the situation of Polish mothers looks pretty good, judging based on the <a href="https://www.oecd.org/els/soc/PF2_1_Parental_leave_systems.pdf" target="_blank">data</a> from 2016.

## Battle Royale: Time vs. Money

Just like with any other area of life, it's also about the trade-off between money and time. The longer you stay with your kid the less money you get. So what would you prefer?

Inspired by the <a href="https://blog.datawrapper.de/weekly37/" target="_blank">visualization</a> from Datawrapper's Lisa Rost, I analyzed the 2016 OECD data on the maternity and parental leave available to mothers.

### Key takeaways

* OECD 2016 **total paid leave average is 1 year 2 weeks** and **payment rate 61%** of the salary mother's salary
* countries of the former Eastern block offer the longest leave
* 3 countries with **the longest leave** are: **Estonia** 166 weeks paid 51%, equivalent to 85 weeks paid 100%, **Slovakia** 164 weeks paid 33% (54 weeks 100%), **Hungary** 160 weeks paid 45% (72 weeks 100%)
* **United States** are the only OECD country that doesn't offer any sort of leave
* **Poland** offers pretty good benefits, reaching almost OECD average (2 weeks less) when it comes to a leave duration and offering 20% more money than OECD average.

Find out more about the different countries benefits by hovering on the data points in the plot. Labels for some countries are hidden to improve readability.

<br>

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, titdy = TRUE,
                      tidy.opts = list(comment = FALSE))
```

```{r}
knitr::read_chunk("parental-leave/parental-leave-oecd.R")
```
```{r plot_leave_vs_paid_rate}

```

<br>

### Note

State and local government can provide additional entitlements which are not included in the analyzed data. Provided data also doesn't account for top-up payments made by employers. Different countries have different taxation rules and social security contributions, which are not reflected in the dataset.

Plotly chart code is available on <a href="https://github.com/olgamie/olga_blog/blob/master/content/post/parental-leave/parental-leave-oecd.R" target="_blank">GitHub</a>.
