---
title: "Building first baseline with random forest using ranger and DALEX"
author: "Olga Mierzwa-Sulima"
date: '2020-09-30'
tags:
- random forest
- variable importance
- DALEX
- ranger
- r
categories: data-science
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# TL;DR
I show an approach to building a first baseline model using random forest (RF).
I use `DALEX` package to understand a random forest built with `ranger` package, simplify it, improve performance, and run time. I use `recipes` package to handle categorical variables, missing values, and novel levels.

## Random forest and first baseline
Random forest is a powerful and flexible model that is relatively easy to tune. This is why it's usually a top candidate to start with and build the first baseline. Establishing a baseline as soon as possible when modeling is a must otherwise if you run a new experiment you don't know if you manage to improve or not.
When starting with RF is not a good idea? Unless you have the data with a very strong time component RF is not a good idea as it [can't extrapolate](https://neptune.ai/blog/random-forest-regression-when-does-it-fail-and-why).

## Variable importance

Feature or variable importance let's you understand which features are important in a given machine learning model. Usually there is only a handful of variables that are important and you want to care about when building and understanding a model.

The main idea behind variable importance is to measure how much the model fit decreases if the effect of a variable is removed. The effect is removed by permuting the column. Values still have the same distribution but it's relation to the dependent variable is gone. If a variable is important, then after random shuffling the model performance should decrease. The larger drop in the performance, the more important is the variable. Read more about pros and cons of this approach in the [Explanatory Model Analysis. Explore, Explain and Examine Predictive Models e-book](https://pbiecek.github.io/ema/featureImportance.html).

`DALEX` package in R offers this powerful and model agnostic way of calculating variable importance.

### How does it affect model performance?
Dropping unimportant columns should not decrease the model performance. By removing a bunch of columns we can get rid of the source of collinearity between the features. The importance that previously was split between the columns should only be reflected in the one left in the model. This makes the feature importance values more clear and reliable.

For example random forest has less columns to try making a split, so it has a slightly better chance to make a better decisions. Model should take less time to run.

The biggest gain comes in less variables that a modeler needs to understand and worry in a model.

## Analysis approach

The main idea is to build a RF model as soon as possible without thinking too much about the data and use model interpretation techniques to make it better. Image that you get to build a model in the domain you know little about, so it's difficult to judge which features might play an important role. Turns out you can use modelling to find out which ones are important and only focus on them in the later stages of analysis.
I use [Blue Book for Bulldozers](https://www.kaggle.com/c/bluebook-for-bulldozers/data) data from Kaggle's  and predict an auction price for a heavy equipment with a random forest. 
I build a model on a data which I pre-process in a very limited way using `recipes` package, just so `ranger` package can actually works. 
After I have the model I plug it into `DALEX` explainer, perform variable importance shuffling and visualize the results. Next by interpreting the results I make adjustments to the 
Example is inspired by [fast.ai](https://www.fast.ai/) analysis from the book [Deep Learning for Coders](https://github.com/fastai/fastbook) chapter on tabular learner's. I show how to do a similar analysis using tools available in R.

## Solution in R

Read in R packages.

```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(ranger)
library(DALEX)
library(tidymodels)
library(rlang)
```

Read in data.

```{r message = FALSE, warning = FALSE}
data_raw <- readr::read_csv(file = "data/Train.csv")
```

### Data pre-processing

There are few things to be done with the data before pluging them into the model.

* I need to handle missing data both in numeric and cardinal columns. In the numeric columns missing values are replaced with median with a `recipe` step `step_medianimpute()`. I add a dummy column for every numeric column that contains NAs using `add_dummy_for_na_numeric()` that I define myself.
I replace NAs in character columns with "NA" string and threat missing values as a separate level.
* I make sure my cardinal variables enter the model as factors. I use `recipe` `step_novel()` to make sure model can handle unseen levels in the validation set.
* I add new features based on sale date column to help the model handle time better. New features are: elapsed number of days since 1 Jan 1970, day, week, month, year, day of week, day of year and indicators is year/quarter/month beginning/end and are added using `proc_date()`.

```{r}
add_dummy_for_na_numeric <- function(x){
  x %>%
    mutate_if(~any(is.na(.)) & is.numeric(.),
              .funs = list(na = ~if_else(is.na(.), 1, 0)))
}

proc_date <- function(data, var_name) {
  data %>%
    dplyr::mutate(
      elapsed = floor(unclass(as.POSIXct({{var_name}}))/86400),
      year = lubridate::year({{var_name}}),
      month = lubridate::month({{var_name}}),
      day = lubridate::day({{var_name}}),
      week = lubridate::week({{var_name}}),
      day_of_week = lubridate::wday({{var_name}}),
      day_of_year = lubridate::yday({{var_name}}),
      is_end_of_month = if_else({{var_name}} == (ceiling_date({{var_name}}, unit = "month") - 1), TRUE, FALSE),
      is_begining_of_month = if_else({{var_name}} == floor_date({{var_name}}, unit = "month"), TRUE, FALSE),
      is_end_of_quarter = if_else({{var_name}} == (ceiling_date({{var_name}}, unit = "quarter") - 1), TRUE, FALSE),
      is_begining_of_quarter = if_else({{var_name}} == floor_date({{var_name}}, unit = "quarter"), TRUE, FALSE),
      is_end_of_year = if_else({{var_name}} == (ceiling_date({{var_name}}, unit = "year") - 1), TRUE, FALSE),
      is_begining_of_year = if_else({{var_name}} == floor_date({{var_name}}, unit = "year"), TRUE, FALSE)
      
    )
    
}
```

Pre-process data by calling helper functions:

```{r}
data_after_basic_preprocess <- data_raw %>%
  dplyr::mutate(
    saledate = lubridate::as_date(saledate, format = "%m/%d/%Y %H:%M")
    ) %>%
  arrange(saledate) %>% 
  proc_date(saledate) %>% 
  mutate_if(is_character, ~if_else(is.na(.), "NA", .)) %>% 
  add_dummy_for_na_numeric()
```
  
I take the log of `SalePrice`. Taking the log of the `SalePrice` makes the variable normally distributed and changes the relative distance between numbers.

```{r}
data_step1 <- data_after_basic_preprocess %>%
  mutate(log_SalePrice = log(SalePrice),
         saledate = NULL,
         SalePrice = NULL)
```

Split data into training and validation set using `initial_time_split()` function. I keep 80% of data in the training set and 20% in the validation. `initial_time_split()` is called on data which is sorted based on `saledate` so I make sure validation set contains "future sales". Taking a random sample would be cheating here, as we would use "future information" to predict the past.

```{r}
data_step1_split <- initial_time_split(data = data_step1, prop = 0.80)

train_set <- training(data_step1_split)

valid_set <- testing(data_step1_split)
```

The rest of the pre-processing is done using `recipes` package:

* Replacing NAs in numeric columns with their median with `step_medianimpute()`
* Changing characters to factors with `step_string2factor()`

Additionally `recipe()` takes care of new levels in validation set using `step_novel()`.

```{r}
recipe_formula <-  recipe(log_SalePrice ~ ., data = data_step1_split)

recipe_steps <- recipe_formula %>%
  step_novel(all_nominal()) %>% 
  step_medianimpute(all_numeric()) %>%
  step_string2factor(all_nominal())

recipe_steps_on_training <- recipe_steps %>% 
  prep(train_set)
```  

Now let's `juice` and `bake`  training and validation sets and split `x` and `y` for `ranger` model to avoid using formula.

```{r}
train_set_preped <- recipe_steps_on_training %>%
  juice()

y_train <- train_set_preped$log_SalePrice
x_train <- train_set_preped %>% select(-log_SalePrice)


valid_set_preped <- recipe_steps_on_training %>%
  bake(valid_set)

x_valid <- valid_set_preped %>% select(-log_SalePrice)
y_valid <- valid_set_preped$log_SalePrice
```


### Model building

I build random forest using `ranger` package. I'm not using `randomForest`, because it has very slow implementation in R.
I limit the number of trees to speed the training process to 40.

To avoid over-fitting:

* use a subset of columns when building trees by setting `mtry` parameter to build less correlated trees,
* use a subset of rows when building trees by setting `sample.fraction`, can be used to build less correlated trees and speed up training,
* stop training the tree further when a leaf node has 3 or less samples. Numbers that usually works well 1, 3, 5, 10, 25.

Categorical variables enter the model as factors and ranger will treat them as ordered. This might seem controversial, but it has been [shown](https://peerj.com/articles/6339/) by the one of the `ranger` authors Marvin Wrigth and Inke König that treating factors like that shouldn't impact final result, but is computationally more efficient. 

```{r}
system.time(ranger_model_with_sampling <- ranger(x = x_train,
                       y = y_train,
                       num.trees = 40,
                       mtry = (ncol(x_train) / 2), 
                       seed = 1,
                       min.node.size = 3,
                       sample.fraction = 1,
                       respect.unordered.factors = "ignore"
                       ))

ranger_predictions <- predict(ranger_model_with_sampling, x_valid)

rmse <- function(preds, y){
  sqrt(mean((preds - y)**2))
}
eval_model <- function(model, y, y_valid, model_predictions) {
  c(model$r.squared,
      rmse(model$predictions, y),
      rmse(model_predictions, y_valid)
  )
}

eval_model(
  ranger_model_with_sampling, y_train, y_valid, ranger_predictions$predictions)
```

### Model interpretation

Now when I have my random forest built I can try to understand it and look at variable importance plot. I use `DALEX` package and create `ranger` explainer. `DALEX` package provides tones of useful tools for understanding black box models.

```{r}
explainer_ranger  <- explain(ranger_model_with_sampling,
                             data = x_train,
                             y = y_train)
```


I use `variable_importance()` function from `ingredients` (DALEX family) package to calculate the variable importance. My loss function is RMSE. DALEX performs by default 10 permutation rounds. I will average the permutation results and inspect the plot for the best 30 features (for better plot readability).
I drop `_baseline_` from the plot. Baseline is the change in model performance when all variables are permuted.

```{r}
var_importance <- variable_importance(explainer_ranger,
                                      loss = loss_root_mean_square,
                                      B = 10,
                                      type = "raw")

summarize_vi <- var_importance %>%
  group_by(variable) %>% 
  summarize(mean_dropout_loss = mean(dropout_loss), .groups = "drop") %>%
  arrange(-mean_dropout_loss) %>%
  mutate(variable = factor(variable, levels = variable))  

ggplot(data = summarize_vi[2:31,],
       aes(y = mean_dropout_loss, x = variable, group = 1)) +
  geom_line() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

There are only few variables that my random forest cares about. `YearMade` is the most important one.
So next I select the features before the curve flattens almost completely at around 0.1 mean dropout loss and build a model on the remaining features subset.

```{r}
var_to_keep <-  summarize_vi %>%
  filter(mean_dropout_loss > 0.1 & variable != "_baseline_") %>% 
  select(variable)

x_after_var_importance <- x_train %>% select(var_to_keep$variable)
```

```{r}
system.time(ranger_model_after_vi <- ranger(x = x_after_var_importance,
                       y = y_train,
                       num.trees = 40,
                       mtry = (ncol(x_after_var_importance) / 2),
                       seed = 1,
                       respect.unordered.factors = F,
                       min.node.size = 3,
                       sample.fraction = 1))

ranger_predictions_after_vi <- predict(ranger_model_after_vi,
                                       x_valid %>% select(var_to_keep$variable))

eval_model(
  ranger_model_after_vi, y_train, y_valid, ranger_predictions_after_vi$predictions)

```

I end up with a simpler model with almost the same quality, differences in R^2 and RMSE are neglectable. I prefer to continue with model with less features which runs faster.

Let's again look at variable importance plot and do more digging.

```{r}
explainer_ranger_after_vi  <- explain(ranger_model_after_vi,
                                      data = x_after_var_importance,
                                      y = y_train)

var_importance_ranger_after_vi <- variable_importance(
  explainer_ranger_after_vi,
  loss_function = loss_root_mean_square,
  B = 10,
  type = "raw")

plot(var_importance_ranger_after_vi)

```

`YearMade` is the most important feature for my model. It would be interesting to learn more about this variable.
Second thing that stands out is how important are SalesID and MachineId. I expect these are unique identifiers so it's strange they are so important. So it's another thing worth looking into.

```{r}
ggplot(data_step1, aes(x = YearMade)) + 
  geom_histogram(binwidth = 10)

data_step1 %>% group_by(YearMade) %>% count()
```

After plotting YearMade distribution I can see that sth weird is happening. There is 38185 entries with YearMade 1000. This is not possible, so I remove this data and use the sales after the War ended.

```{r}
data_step1$SalesID %>% n_distinct()
data_step1$MachineID %>% n_distinct()
data_step1 %>%
  filter(MachineID == 2283592)
```
SalesID is a unique identifier that sorts sales from the oldest to the newest.
Kaggle's description says MachineID is unique machine identifier and that machine can be sold multiple times. I expect that RF model could be overfitting on these columns so I remove them from the dataset and see how it impacts the model.

```{r}
x_train <- x_train %>%
  mutate(YearMade = if_else(YearMade >= 1950, YearMade, 1950)) %>%
  select(var_to_keep$variable) %>%
  select(-c(SalesID, MachineID))

x_valid <- x_valid %>%
  mutate(YearMade = if_else(YearMade >= 1950, YearMade, 1950)) %>%
  select(var_to_keep$variable) %>%
  select(-c(SalesID, MachineID))
```

```{r}
system.time(ranger_model_after_vi_data_clean <- ranger(x = x_train,
                       y = y_train,
                       num.trees = 40,
                       mtry = (ncol(x_train) / 2),
                       seed = 1,
                       min.node.size = 3,
                       sample.fraction = 1))

ranger_predictions_after_vi_clean <- predict(ranger_model_after_vi_data_clean,
                                       x_valid)

eval_model(
  ranger_model_after_vi_data_clean, y_train, y_valid, ranger_predictions_after_vi_clean$predictions)
```

It seems cleaning YearMade and dropping IDs improved the model just a bit.

## Conclusions

I manage to fairly quick build a first model that can serve as a baseline in the future experiments. I used variable importance analysis to simplify the model and get the first insights into features that matter for the model. This knowledge could be further used to do more sophisticated feature engineering. 
